Basics of Hardware Cache Optimization

### 2017-05-30 tex/basic-concepts/lines.tex
Add new measurement; revise, expand, and tweak

My E-450 CPU is no exception and both of its data caches have 64-byte cache lines.%
\footnote{See \cref{app:cpuinfo}.}
% \begin{minted}[gobble=3]{bash}
%    $ getconf LEVEL1_DCACHE_LINESIZE; getconf LEVEL2_CACHE_LINESIZE
%    64
%    64
% \end{minted}
%stopzone
% \footnote{%
%    \mintinline{bash}!$ getconf LEVEL1_DCACHE_LINESIZE; getconf LEVEL2_CACHE_LINESIZE!\\
%    %stopzone
%    \noindent\mintinline{bash}!64!\\
%    \noindent\mintinline{bash}!64!
% }
We can verify this quite easily.  Consider \cref{lst:line-size}.  It loops over an array
with an increment given at compile time as \texttt{STEP} and measures the processor time.
\begin{center}
   \inputminted[firstline=27]{c}{line-size/line-size.c}
   \captionof{listing}{%
      Loop over \mintinline{text}{array} with increment \mintinline{text}{STEP}}
   \label{lst:line-size}
\end{center}
The results for different values of \texttt{STEP} are plotted in \cref{fig:line-size}.
% Starting from a step size of 16, the time roughly halves every time the step size is
% doubled.  For the first 4 step sizes however, it is almost constant.
As expected, the time roughly halves whenever the step size is doubled --- but only from a
step size of 16.  For the first 4 step sizes, it is almost constant.

@@

\begin{comment}
   \begin{listing}
      \inputminted[firstline=27]{c}{line-size/line-size.c}
      \caption{Loop over \mintinline{text}{array} with increment \mintinline{text}{STEP}}
      \label{lst:line-size}
   \end{listing}
\end{comment}

\begin{figure}
   \centering
   \begin{tikzpicture}
      \datavisualization
      [scientific axes=clean,
       x axis={logarithmic,
               ticks={major={at={1 as \textbf{1}, 2 as \textbf{2}, 4 as \textbf{4},
                                 8 as \textbf{8}, 16, 64, 256, 1024}},
                      minor={at={32, 128, 512}}},
               label={Step size}, length=0.8\textwidth},
       y axis={logarithmic,
               % ticks={major={at={4, 16, 64, 256}}, minor={at={8, 32, 128}}},
               % ticks={major={at={6, 12, 24, 48, 92, 184}}},
               ticks={major={at={5, 10, 20, 40, 80, 160, 320}}},
               label={Processor time (ms)}, length=6cm},
       visualize as scatter,
       scatter={style={mark=*, mark options={scale=.65}}}]
         % See <https://tex.stackexchange.com/q/198323>.
         data [read from file=line-size/line-size.csv, separator=\space];
   \end{tikzpicture}
   \caption{Processor times for running \cref{lst:line-size}}
   \label{fig:line-size}
\end{figure}

@@

@@

>>>>

My E-450 CPU is no exception and both of its data caches have 64-byte cache lines.%
\footnote{See \cref{app:cpuinfo}.}
% \begin{minted}[gobble=3]{bash}
%    $ getconf LEVEL1_DCACHE_LINESIZE; getconf LEVEL2_CACHE_LINESIZE
%    64
%    64
% \end{minted}
%stopzone
% \footnote{%
%    \mintinline{bash}!$ getconf LEVEL1_DCACHE_LINESIZE; getconf LEVEL2_CACHE_LINESIZE!\\
%    %stopzone
%    \noindent\mintinline{bash}!64!\\
%    \noindent\mintinline{bash}!64!
% }
We can verify this quite easily.  Consider \cref{lst:line-size}.  It loops over an array
with an increment given at compile time as \texttt{STEP} and measures the processor time.
\begin{center}
% \begin{listing}
   \inputminted[firstline=27]{c}{line-size/line-size.c}
   % \caption{%
   \captionof{listing}{%
      Loop over \mintinline{text}{array} with increment \mintinline{text}{STEP}}
   \label{lst:line-size}
% \end{listing}
\end{center}
The results for different values of \texttt{STEP} are plotted in \cref{fig:line-size}.
% Starting from a step size of 16, the time roughly halves every time the step size is
% doubled.  For the first 4 step sizes however, it is almost constant.
As expected, the time roughly halves whenever the step size is doubled --- but only from a
step size of 16.  For the first 4 step sizes, it is almost constant.

@@

@@

% TODO.  Part of the memory address partially determines which cache line is used.  The
% binary logarithm of the number of cache sets is the amount of bits required to identify
% the cache set.  The specific line used from that set depends on the replacement policy.
% A number of least significant (lowest-order, right-most) bits are ignored, since all
% addresses that only differ in those go into the same cache line.  If the line size is
% 64, log_2(64) = 6 bits are ignored.  See Drepper, page 15.

@@

% XXX: hacks!  Explicit placement specifier without `t` (top) to prevent the figure from
% interrupting the source code listing.
\begin{figure}[hbp]
   \centering
   \begin{tikzpicture}
      \datavisualization
      [scientific axes=clean,
       x axis={logarithmic,
               ticks={major={at={1 as \textbf{1}, 2 as \textbf{2}, 4 as \textbf{4},
                                 8 as \textbf{8}, 16, 64, 256, 1024}},
                      minor={at={32, 128, 512}}},
               label={Step size}, length=0.8\textwidth},
       y axis={logarithmic,
               % ticks={major={at={4, 16, 64, 256}}, minor={at={8, 32, 128}}},
               % ticks={major={at={6, 12, 24, 48, 92, 184}}},
               ticks={major={at={10, 20, 40, 80, 160, 320}}},
               label={Processor time (ms)}, length=6cm},
       visualize as scatter,
       scatter={style={mark=*, mark options={scale=.65}}}]
         % See <https://tex.stackexchange.com/q/198323>.
         data [read from file=line-size/line-size.csv, separator=\space];
   \end{tikzpicture}
   \caption{Processor times for running \cref{lst:line-size}}
   \label{fig:line-size}
\end{figure}

### 2017-05-30 tex/basic-concepts/prefetching.tex
Add new measurement; revise, expand, and tweak

\subsection{Prefetching}
\label{sec:prefetch}
% http://ithare.com/c-for-games-performance-allocations-and-data-locality/
% http://ithare.com/c-for-games-performance-allocations-and-data-locality/2/
% https://en.wikipedia.org/wiki/Cache_prefetching

\emph{Cache prefetching} is a technique by which CPUs recognize predictable access
patterns (for instance sequential access to an array)... TODO

>>>>

\subsection{Prefetching}
\label{sec:prefetch}

% \cref{lst:TODO} is
Consider
a simplified version of \cref{lst:access-times} that, instead of using random accesses,
simply walks over the array sequentially.  It still follows the pointers to do this, but
the array is no longer shuffled.

% \emph{Cache prefetching} is a technique by which CPUs recognize predictable access
% patterns (for instance sequential access to an array)... TODO

% [1]: http://ithare.com/c-for-games-performance-allocations-and-data-locality/
% [2]: http://ithare.com/c-for-games-performance-allocations-and-data-locality/2/
% [3]: https://en.wikipedia.org/wiki/Cache_prefetching

### 2017-05-30 tex/types.tex
Add new measurement; revise, expand, and tweak

\begin{center}
   \begin{tabular}{ r | c c c c }
             & \gls{l1d} & \gls{l2} & \gls{l3} & Main Memory \\ \hline
      Cycles & 3--4      & 10--12   & 30--70   & 100--150
   \end{tabular}
\end{center}
%
% These are taken from~\textcite{ithare-cycles} but comparable numbers are given by

@@

Up to 32 KiB, each access takes almost exactly 3 cycles.%
\footnote{The numerical results are shown in \vref{tab:access-times}.}
This is the \gls{l1d} access \alts{time, latency}.  At 32 KiB (the size of the \gls{l1d})
the time increases to about 3.4 cycles.  This is not surprising since the cache is shared
with other processes and the operating system, so some of our data gets \emph{evicted}.
The first dramatic increase happens at 64 KiB followed by smaller increases at 128 and 256
KiB.  I suspect we are seeing a mixture of \gls{l2} and \gls{l1d} accesses, with less and
less \gls{l1d} hits and an \gls{l2} access time \x{of} around 25 cycles.

@@

% XXX: hacks!  Explicit placement specifier without `t` (top) to prevent the figure from
% interrupting the source code listing.
\begin{figure}[hbp]
   \centering
   \begin{tikzpicture}
      \datavisualization[%
         scientific axes=clean,
         x axis={%
            logarithmic,
            ticks={major={
               /pgf/number format/int detect,
               at={
                  2048 as 2, 8192 as 8, 32768 as \textbf{32}, 131072 as 128,
                  524288 as \textbf{512}, 2097152 as \pgfmathprintnumber{2048},
                  8388608 as \pgfmathprintnumber{8192},
                  33554432 as \pgfmathprintnumber{32768},
                  134217728 as \pgfmathprintnumber{131072},
                  % $128\cdot 2^{10}$}}},
                  % $2^{17}$}}},
            }}},
            grid={at={32768, 524288}},
            label={Array Size (KiB)},
            length=0.8\textwidth},
         y axis={include value=0, label={Cycles / Iteration}, length=6cm, grid=at ticks},
         visualize as scatter,
         scatter={style={mark=*, mark options={scale=.65}}}]
         data [read from file=access-times/access-times.csv, separator=\space];
   \end{tikzpicture}
   % \caption{Access Times for Random Reads (\Cref{lst:access-times})}
   \caption{Access Times for Random Reads}
   \label{fig:access-times}
\end{figure}

>>>>

% \begin{center}
%    \begin{tabular}{ r | c c c c }
%              & \gls{l1d} & \gls{l2} & \gls{l3} & Main Memory \\ \hline
%       Cycles & 3--4      & 10--12   & 30--70   & 100--150
%    \end{tabular}
% \end{center}
\begin{center}\begin{tabular}{ r c c c c }
   \toprule
          & \gls{l1d} & \gls{l2} & \gls{l3} & Main Memory \\
   \cmidrule[\lightrulewidth](l){2-5}
   % \midrule
   Cycles & 3--4      & 10--12   & 30--70   & 100--150 \\
   \bottomrule
\end{tabular}\end{center}

%
% These are taken from~\textcite{ithare-cycles} but comparable numbers are given by

@@

Up to 32 KiB, each access takes almost exactly 3 cycles.%
\footnote{The numerical results are shown in \vref{tab:access-times}.}
This is the \gls{l1d} access \alts{time, latency}.  At 32 KiB (the size of the \gls{l1d})
the time increases to about 3.4 cycles.  This is not surprising since the cache is shared
with other processes and the operating system, so some of our data gets \emph{evicted}.
The first dramatic increase happens at 64 KiB followed by smaller increases at 128 and 256
KiB.  I suspect we are seeing a mixture of \gls{l2} and \gls{l1d} accesses, with less and
less \gls{l1d} \emph{hits} and an \gls{l2} access time \x{of} around 25 cycles.

@@

% XXX: hacks!  Explicit placement specifier without `t` (top) to prevent the figure from
% interrupting the source code listing.
\begin{figure}%[hbp]
   \centering
   \begin{tikzpicture}
      \datavisualization[%
         scientific axes=clean,
         x axis={%
            logarithmic,
            ticks={major={
               /pgf/number format/int detect,
               at={
                  2048 as 2, 8192 as 8, 32768 as \textbf{32}, 131072 as 128,
                  524288 as \textbf{512}, 2097152 as \pgfmathprintnumber{2048},
                  8388608 as \pgfmathprintnumber{8192},
                  33554432 as \pgfmathprintnumber{32768},
                  134217728 as \pgfmathprintnumber{131072},
                  % $128\cdot 2^{10}$}}},
                  % $2^{17}$}}},
            }}},
            grid={at={32768, 524288}},
            label={Array Size (KiB)},
            length=0.8\textwidth},
         y axis={include value=0, label={Cycles / Iteration}, length=6cm, grid=at ticks},
         visualize as scatter,
         scatter={style={mark=*, mark options={scale=.65}}}]
         data [read from file=access-times/access-times.csv, separator=\space];
   \end{tikzpicture}
   % \caption{Access Times for Random Reads (\Cref{lst:access-times})}
   \caption{Access Times for Random Reads}
   \label{fig:access-times}
\end{figure}
